`traveltime`: an R package to calculate travel time across a landscape from user-specified locations

### 2025-04-09 paper/paper.qmd
flesh out again, render to docx

---
title: "`traveltime`: an R package to calculate travel time across a landscape from user-specified locations"
format:
  # md: default
  # wordcount-html:
  # html:
  #   keep-md: true
  #   fig-height: 4
  #   fig-align: center
  #   fig-format: png
  #   dpi: 300
  pdf:
    keep-md: true
    fig-height: 4
    fig-align: center
    dpi: 300
    fig-format: png
header-includes:
  \usepackage{lineno} \linenumbers
  \usepackage{hanging}
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 4"
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
 - name: University of Western Australia, WA, Australia
   index: 4
date: 2025-04-07
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of geographic coordinates. Applications include: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].

@@

# Statement of need

@@

A gaggle of R packages provide superficially similar though fundamentally different functionality via the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. Their 'Isochron' polygons --- areas reachable within a given time from a given location --- are most comparable to what `traveltime::calculate_travel_time()` calculates. However, each isochron is a single polygon calculated is for a single point location and specified maximum travel time, rather than a raster of temporal gradation across a landscape, jointly for an arbitrary number of points, as in `traveltime`.

@@

Resulting travel times are Figure \ref{fig-result}[^2].

@@

[^2]: In Figure \ref{fig-data}, the islands to the south and north-east are filled, but unconnected with the mainland. These cells appear absent in Figure \ref{fig-result}; because they are not connected to any cells with a station, the calculated travel time is infinite, and so are not plotted in Figure \ref{fig-result}.

@@

>>>>

---
title: "`traveltime`: an R package to calculate travel time across a landscape from user-specified locations"
format:
  # md: default
  # wordcount-html:
  # html:
  #   keep-md: true
  #   fig-height: 4
  #   fig-align: center
  #   fig-format: png
  #   dpi: 300
  docx:
    keep-md: true
    fig-height: 4
    fig-align: center
    dpi: 300
    fig-format: png
header-includes:
  \usepackage{lineno} \linenumbers
  \usepackage{hanging}
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 4"
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
 - name: University of Western Australia, WA, Australia
   index: 4
date: 2025-04-07
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a raster map of the travel time over an area of interest from a user-specified set of locations (geographic coordinates). The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to the nearest of the supplied locations. We envisage this software having diverse applications including: estimating sampling bias [@dennis2000bias], allocating defibrillators [@tierney2018novel], setting health districts [@padgham2019introduction], or mapping access to vehicle chargers [@falchetta2021electric] and agricultural facilities [@zhao2023replanting].

@@

# Context

@@

A gaggle of R packages provide superficially similar though fundamentally different functionality via the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. Their 'Isochron' polygons --- areas reachable within a given time from a given location --- are most comparable to what `traveltime::calculate_travel_time()` calculates. However, each isochron is a single polygon calculated is for a single point location and specified maximum travel time, rather than a raster of temporal gradation across a landscape, jointly for an arbitrary number of points, as in `traveltime`. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

@@

We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away. Note that the results in `trave_time_singapore` include infinite values (`Inf` above). In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland. The raster cells for these islands appear absent in Figure \ref{fig-result}. Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.

@@

@@

Firstly, `traveltime` currently facilitates access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2]. Although the user can presently supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global]. As landscapes are dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell. We underline however that a user can provide their own higher resolution friction surface at present if desired.

[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
